Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-23 RevisedSubmission/hines_hetland_2017_gji.tex
added prior demo

We seek a spatially and temporally dependent estimate of transient
strain rates. We consider transient strain rates to be any deviation
from secular strain rates. We denote transient strain rates as
$\strain(p)$, where $p$ represents the ordered pair $(\pos,t)$, $\pos
= (x_\e,x_\n)$ are spatial coordinates, and $t$ is time. For
simplicity, we assume that the study region is sufficiently small that
$\pos$ can be considered a point in a 2-D Cartesian map projection
that is aligned with the cardinal directions. We determine $\strain$
by spatially and temporally differentiating estimates of transient
displacements (i.e., displacements that deviate from the secular
trend), which are constrained by GNSS data.

@@

We provide an example to give our prior on $\vec{u}$ a more tangible
meaning. we can use a squared exponential function for $X_i$ and
$T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2\ell^2}\right), \ \ \ \ 
T_i(t,t') = \phi^2 \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
which satisfies our requirements for positive definiteness and
differentiability. The parameters $\ell$ and $\tau$ control the
length-scale and time-scale, respectively, of realizations (i.e.,
sample functions) of $u_i$. The parameter $\phi$, which we have
arbitrarily chosen to incorporate into $T_i$ rather than $X_i$,
controls the amplitude of realizations of $u_i$. Ideally, we want
realizations of $u_i$ to have a length-scale, time-scale, and
amplitude that resembles what we expect from true signal. A few
realizations for $u_i$ are shown in Figure XXX. While the squared
exponential function is a commonly used covariance function for GPR,
it is not appropriate for every application. The appropriate choice
for $X_i$ and $T_i$ may vary depending on the geophysical signal we
are trying to describe. To keep this section sufficiently general, we
hold off on specifying $X_i$ and $T_i$ until Section
\ref{sec:SignalModel}, where we estimate transient strain from slow
slip events in the Pacific Northwest.

@@

The posterior transient displacements are spatially and temporally
continuous, and we can use eqs. (\ref{eq:PosteriorMean3}) and
(\ref{eq:PosteriorCov3}) to evaluate $\post_i$ at any $p$.
Furthermore, $\post_i$ is spatially and temporally differentiable,
allowing us to formulate $\strain$ at any $p$ that we may be
interested in. The components of $\strain$ can be written as
\begin{equation}\label{eq:StrainRate}
\strain_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} 
                  \left(\frac{\partial \post_i(p)}{\partial x_j} +  
                        \frac{\partial \post_j(p)}{\partial x_i} \right).
\end{equation}
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian
processes $\post_i$ and $\post_j$, we know that $\strain_{ij}$ is also
a Gaussian process. From \citet[sec. 10]{Papoulis1991}, we find that
the mean and covariance functions for $\dot{\varepsilon}_{ij}$ are
\begin{align}\label{eq:StrainMean}
\mu_{\strain_\e\e}(p) &= \frac{\partial^2 \mu_{\post_\e}(p)}{\partial t \, \partial x_\e} \\
\mu_{\strain_\n\n}(p) &= \frac{\partial^2 \mu_{\post_\n}(p)}{\partial t \, \partial x_\n} \\
\mu_{\strain_\e\n}(p) &= \frac{1}{2}\frac{\partial}{\partial t}\left(
                                    \frac{\partial \mu_{\post_\e}(p)}{\partial x_\n} + 
                                    \frac{\partial \mu_{\post_\n}(p)}{\partial x_\e} \right)
\end{align} 
and
\begin{align}\label{eq:StrainCov}
C_{\strain_\e\e}(p,p') &= \frac{\partial^4 C_{\post_\e}(p,p')}{\partial t \, \partial t' \, \partial x_\e \, \partial x'_\e} \\
C_{\strain_\n\n}(p,p') &= \frac{\partial^4 C_{\post_\n}(p,p')}{\partial t \, \partial t' \, \partial x_\n \, \partial x'_\n} \\
C_{\strain_\e\n}(p,p') &= \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                      \frac{\partial^2 C_{\post_\e}(p,p')}{\partial x_\n \, \partial x'_\n} + 
                                      \frac{\partial^2 C_{\post_\n}(p,p')}{\partial x_\e \, \partial x'_\e} \right),
\end{align} 
respectively.

@@

Our motivation for estimating transient strain rates is, in part, to
detect geophysical phenomena. As we will see, geophysical signal can
be easily identified by visually inspecting the solution for $\strain$
from eqs. (\ref{eq:StrainMean}) and (\ref{eq:StrainCov}). However, if
we want to detect geophysical phenomena automatically, then we need to
define a detection criterion. We use a signal-to-noise ratio, SNR,
that is based on the Frobenius norm of $\strain$,
$||\strain||_\mathrm{F} = \left(\strain_{\e\e}^2 + \strain_{\n\n}^2 +
2\strain_{\e\n}^2\right)^{\frac{1}{2}}$, for our detection criterion.
In the geodetic literature, $||\dot{\varepsilon}||_\mathrm{F}$ is
often used as a metric for the strain rate ``magnitude", and it is
sometimes referred to as the second invariant of strain rate. Noting
that $||\strain||_\mathrm{F}$ is a random variable, we take SNR to be
the ratio of the estimated mean and standard deviation of
$||\strain||_\mathrm{F}$. An estimate of the mean is found by
evaluating $||\strain||_\mathrm{F}$ at the mean of $\strain$,
\begin{align}
\mu_{||\strain||_\mathrm{F}} &\approx ||\strain||_F \big|_{\strain = \mu_{\strain}}\nonumber \\
                             &= \left(\mu_{\strain_\e\e}^2 + 
                                      \mu_{\strain_\n\n}^2 + 
                                      2\mu_{\strain_\e\n}^2 \right)^\frac{1}{2},
\end{align}
and we use nonlinear uncertainty propagation to estimate the standard
deviation,
\begin{equation}
\sigma_{||\strain||_\mathrm{F}} \approx
\left( \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_\e\e}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_\e\e} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_\n\n}\right|_{\strain = \mu_{\strain}} \right)^2
       \sigma^2_{\strain_\n\n} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_\e\n}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_\e\n} \right)^\frac{1}{2},
\end{equation}
where $\sigma^2_{\strain_{ij}}(p) = C_{\strain_{ij}}(p,p)$. After some
calculations, we find SNR to be
\begin{align}\label{eq:SNR}
\mathrm{SNR}(p) &= \frac{\mu_{||\strain||_\mathrm{F}}(p)}{\sigma_{||\strain||_\mathrm{F}}(p)} \\
                &= \frac{\mu_{\strain_\e\e}(p)^2 +
                         \mu_{\strain_\n\n}(p)^2 +
                         2\mu_{\strain_\e\n}(p)^2}
                        {\big(\sigma^2_{\strain_\e\e}(p)\mu_{\strain_\e\e}(p)^2 + 
                              \sigma^2_{\strain_\n\n}(p)\mu_{\strain_\n\n}(p)^2 + 
                              4\sigma^2_{\strain_\e\n}(p)\mu_{\strain_\e\n}(p)^2
                         \big)^{\frac{1}{2}}}.
\end{align}
We explicitly show that SNR is a function of $p$ to emphasize that it
identifies the position and time of anomalous deformation. We can
reasonably suspect that some transient geophysical phenomena is
occurring wherever and whenever SNR is larger than ${\sim}3$.

>>>>

We seek a spatially and temporally dependent estimate of transient
strain rates. We consider transient strain rates to be any deviation
from the secular strain rates. We denote transient strain rates as
$\strain(p)$, where $p$ represents the ordered pair $(\pos,t)$, $\pos
= (x_\e,x_\n)$ are spatial coordinates, and $t$ is time. For
simplicity, we assume that the study region is sufficiently small that
$\pos$ can be considered a point in a 2-D Cartesian map projection
that is aligned with the cardinal directions. We determine $\strain$
by spatially and temporally differentiating estimates of transient
displacements, which are constrained by GNSS data.

@@

% Example of the prior

We provide an example to give our prior on $\vec{u}$ a more tangible
meaning. we can use a squared exponential function for $X_i$ and
$T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2\ell^2}\right), \ \ \ \ 
T_i(t,t') = \phi^2 \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
which satisfies our requirements for positive definiteness and
differentiability. The parameters $\ell$ and $\tau$ control the
length-scale and time-scale, respectively, of realizations (i.e.,
sample functions) of $u_i$. The parameter $\phi$, which we have
arbitrarily chosen to incorporate into $T_i$ rather than $X_i$,
controls the amplitude of realizations of $u_i$. Ideally, we want
realizations of $u_i$ to have a length-scale, time-scale, and
amplitude that resembles what we expect for the true transient
displacements. A few realizations for $u_i$ are shown in Figure XXX.
While the squared exponential function is a commonly used covariance
function for GPR, it is not appropriate for every application. The
appropriate choice for $X_i$ and $T_i$ may vary depending on the
geophysical signal we are trying to describe. To keep this section
sufficiently general, we hold off on specifying $X_i$ and $T_i$ until
Section \ref{sec:SignalModel}, where we estimate transient strain from
slow slip events in the Pacific Northwest.

@@

The posterior transient displacements are spatially and temporally
continuous, and we can use eqs. (\ref{eq:PosteriorMean3}) and
(\ref{eq:PosteriorCov3}) to evaluate $\post_i$ at any $p$.
Furthermore, $\post_i$ is spatially and temporally differentiable,
allowing us to formulate $\strain$ at any $p$ that we may be
interested in. The components of $\strain$ can be written as
\begin{equation}\label{eq:StrainRate}
\strain_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} 
                  \left(\frac{\partial \post_i(p)}{\partial x_j} +  
                        \frac{\partial \post_j(p)}{\partial x_i} \right).
\end{equation}
Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian
processes $\post_i$ and $\post_j$, we know that $\strain_{ij}$ is also
a Gaussian process. From \citet[sec. 10]{Papoulis1991}, we find that
the mean and covariance functions for $\dot{\varepsilon}_{ij}$ are
\begin{align}\label{eq:StrainMean}
\mu_{\strain_{\e\e}}(p) &= \frac{\partial^2 \mu_{\post_\e}(p)}{\partial t \, \partial x_\e} \\
\mu_{\strain_{\n\n}}(p) &= \frac{\partial^2 \mu_{\post_\n}(p)}{\partial t \, \partial x_\n} \\
\mu_{\strain_{\e\n}}(p) &= \frac{1}{2}\frac{\partial}{\partial t}\left(
                                    \frac{\partial \mu_{\post_\e}(p)}{\partial x_\n} + 
                                    \frac{\partial \mu_{\post_\n}(p)}{\partial x_\e} \right)
\end{align} 
and
\begin{align}\label{eq:StrainCov}
C_{\strain_{\e\e}}(p,p') &= \frac{\partial^4 C_{\post_\e}(p,p')}{\partial t \, \partial t' \, \partial x_\e \, \partial x'_\e} \\
C_{\strain_{\n\n}}(p,p') &= \frac{\partial^4 C_{\post_\n}(p,p')}{\partial t \, \partial t' \, \partial x_\n \, \partial x'_\n} \\
C_{\strain_{\e\n}}(p,p') &= \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                      \frac{\partial^2 C_{\post_\e}(p,p')}{\partial x_\n \, \partial x'_\n} + 
                                      \frac{\partial^2 C_{\post_\n}(p,p')}{\partial x_\e \, \partial x'_\e} \right),
\end{align} 
respectively.

@@

Our motivation for estimating transient strain rates is, in part, to
detect geophysical phenomena. As we will see, geophysical signal can
be easily identified by visually inspecting the solution for $\strain$
from eqs. (\ref{eq:StrainMean}) and (\ref{eq:StrainCov}). However, if
we want to detect geophysical phenomena automatically, then we need to
define a detection criterion. We use a signal-to-noise ratio, SNR,
that is based on the Frobenius norm of $\strain$,
$||\strain||_\mathrm{F} = \left(\strain_{\e\e}^2 + \strain_{\n\n}^2 +
2\strain_{\e\n}^2\right)^{\frac{1}{2}}$, for our detection criterion.
In the geodetic literature, $||\dot{\varepsilon}||_\mathrm{F}$ is
often used as a metric for the strain rate ``magnitude", and it is
sometimes referred to as the second invariant of strain rate. Noting
that $||\strain||_\mathrm{F}$ is a random variable, we take SNR to be
the ratio of the estimated mean and standard deviation of
$||\strain||_\mathrm{F}$. An estimate of the mean is found by
evaluating $||\strain||_\mathrm{F}$ at the mean of $\strain$,
\begin{align}
\mu_{||\strain||_\mathrm{F}} &\approx ||\strain||_F \big|_{\strain = \mu_{\strain}}\nonumber \\
                             &= \left(\mu_{\strain_{\e\e}}^2 + 
                                      \mu_{\strain_{\n\n}}^2 + 
                                      2\mu_{\strain_{\e\n}}^2 \right)^\frac{1}{2},
\end{align}
and we use nonlinear uncertainty propagation to estimate the standard
deviation,
\begin{equation}
\sigma_{||\strain||_\mathrm{F}} \approx
\left( \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\e}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\e}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\n\n}}\right|_{\strain = \mu_{\strain}} \right)^2
       \sigma^2_{\strain_{\n\n}} +
       \left(\left. \frac{\partial ||\strain||_\mathrm{F}}{\partial \strain_{\e\n}}\right|_{\strain = \mu_{\strain}} \right)^2 
       \sigma^2_{\strain_{\e\n}} \right)^\frac{1}{2},
\end{equation}
where $\sigma^2_{\strain_{ij}}(p) = C_{\strain_{ij}}(p,p)$. After some
calculations, we find SNR to be
\begin{align}\label{eq:SNR}
\mathrm{SNR}(p) &= \frac{\mu_{||\strain||_\mathrm{F}}(p)}{\sigma_{||\strain||_\mathrm{F}}(p)} \\
                &= \frac{\mu_{\strain_{\e\e}}(p)^2 +
                         \mu_{\strain_{\n\n}}(p)^2 +
                         2\mu_{\strain_{\e\n}}(p)^2}
                        {\big(\sigma^2_{\strain_{\e\e}}(p)\mu_{\strain_{\e\e}}(p)^2 + 
                              \sigma^2_{\strain_{\n\n}}(p)\mu_{\strain_{\n\n}}(p)^2 + 
                              4\sigma^2_{\strain_{\e\n}}(p)\mu_{\strain_{\e\n}}(p)^2
                         \big)^{\frac{1}{2}}}.
\end{align}
We explicitly show that SNR is a function of $p$ to emphasize that it
identifies the position and time of anomalous deformation. We can
reasonably suspect that some transient geophysical phenomena is
occurring wherever and whenever SNR is larger than ${\sim}3$.
